---
title: "Econ Data Exploration"
date: "7 March, 2016"
output:
  html_document:
    code_folding: hide
---

## Introduction

This analysis is a brief exploration of the World Economic Indicators dataset downloaded from http://datacatalog.worldbank.org/. It's to show feasibility that this dataset is actually analyzable and therefore suitable for our Econ Data competition (http://www.slideshare.net/TonyHui3/big-data-big-impact-info-session-slides)

```{r, include=FALSE}
require(data.table)
require(ggplot2)
require(tidyr)
require(knitr)
require(pheatmap)
library(MASS)
require(dplyr)
opts_chunk$set(warning=FALSE, messages=FALSE, cache=TRUE, fig.width = 4, fig.height = 3, cache=TRUE)
```

**Load data**

```{r load_data, cache=TRUE, cache.lazy=TRUE}
all_data <- fread("WDI_Data.csv", header = T, check.names = TRUE, data.table = FALSE) %>%
  setnames(gsub(" ", "", colnames(.)) %>% tolower()) %>% tbl_df()

all_data_long <- all_data %>%
  select(-countryname, -indicatorname) %>%
  gather(key = year, value = value, -countrycode, -indicatorcode) %>%
  filter(!is.na(value))
```

```{r}
DT::datatable(all_data[sample(1:nrow(all_data), size = 10),], style = "bootstrap", rownames = FALSE, fillContainer = TRUE)
```

## Find GDP related indicators

`n` is the number of country/year pairs that data is available for this

```{r}
indicator_codes <- data.frame(indicatorcode = all_data_long$indicatorcode %>% unique, stringsAsFactors = F)

all_data_long %>% 
  filter(grepl("GDP", indicatorcode)) %>%
  group_by(indicatorcode) %>%
  tally() %>%
  arrange(desc(n)) %>%
  head(5) %>%
  kable(format = "html")

# Select GDP per capita growth (annual %)

gdp <- "NY.GDP.PCAP.CD"
```

Will use `r gdp` since it has the most data points

## Format data for correlations

- Remove variables where latest measurement isn't 2012 or later
- Remove variables with less than 10 years of data, or variables with 0 standard deviation
- Only grab datapoints where the data lies between the GDP datapoints
- Results in a matrix, where each row represents one year

```{r}
ldc_data_long <- all_data_long %>% 
  filter(countrycode == "LDC")

# Get earliest and latest timepoint for GDP data, removing the rest
ldc_data_daterange <- ldc_data_long %>%
  filter(indicatorcode == gdp) %>%
  summarize(min = min(year), max = max(year))

ldc_data_daterange

# Get variable names to include 
good_indicators <- ldc_data_long %>% 
  group_by(indicatorcode) %>%
  filter(n() > 10, sd(value) > 0, max(year) >= 2012) %>%
  dplyr::select(`indicatorcode`) %>%
  unique()

# Transform into matrix
ldc_data_matrix <- ldc_data_long %>%
  filter(year >= ldc_data_daterange$min, year <= ldc_data_daterange$max) %>%
  filter(indicatorcode %in% good_indicators$indicatorcode) %>%
  spread(key = indicatorcode, value = value) %>%
  select(-c(1:2))

ldc_data_matrix[1:6,1:6] %>% DT::datatable(fillContainer = TRUE)
```

## Correlate each indicator with `r gdp` of LDC

```{r}
# correlate
ldc_cor_matrix <- cor(x = as.matrix(ldc_data_matrix[,gdp]), y = as.matrix(select(ldc_data_matrix, -matches(gdp))), use = "pairwise.complete", method = "spearman")

ldc_cor_df <- as.data.frame(ldc_cor_matrix) %>% t %>% data.frame() %>%
  tibble::rownames_to_column("indicatorcode") %>%
  rename(LDC = NY.GDP.PCAP.CD) %>%
  arrange(desc(abs(LDC)))

ldc_cor_df %>% head

highest_cor <- ldc_cor_df$indicatorcode[1]
```

## plot this

Don't worry about y axis labels - the point is to show that some are positive and some are negative

```{r, fig.height=5}
ldc_cor_df %>%
  ggplot(aes(reorder(indicatorcode, LDC), LDC)) +
  geom_bar(stat = "identity") +
  coord_flip()
```

One can also represent the height of the bar with color - in this case, higher bar = higher correlation = more red

```{r, fig.height=5}
ldc_cor_df %>%
  ggplot(aes(reorder(indicatorcode, LDC), "LDC", fill=LDC)) +
  geom_bin2d() +
  scale_fill_gradient2(low = "#2c7bb6", mid = "#ffffbf", high = "#d7191c") +
  coord_flip()
```

## "Visually" confirm the highest correlation

[`r highest_cor`](`r paste0("http://data.worldbank.org/indicator/",highest_cor,"?end=2015&start=2003&view=chart")`) is *Cost of business start-up procedures*

```{r}
test <- "IC.REG.COST.PC.ZS"

ldc_confirm <- ldc_data_long %>%
  filter(indicatorcode == gdp | indicatorcode == test) %>%
  spread(key = indicatorcode, value = value) %>%
  filter(complete.cases(.)) %>%
  as.data.frame()

ggplot(ldc_confirm, aes(x = ldc_confirm[,gdp], y = ldc_confirm[,test])) +
  xlab(gdp) +
  ylab(test) +
  geom_point() +
  geom_smooth(method = "lm", se = F)
```

## Scale computation to all countries

```{r scale_to_all, cache=TRUE}
# Get earliest and latest timepoint for GDP data, removing the rest
all_data_daterange <- all_data_long %>%
  group_by(countrycode) %>%
  filter(indicatorcode == gdp) %>%
  summarize(min_year = as.numeric(min(year)), max_year = as.numeric(max(year))) %>%
  mutate(range = max_year - min_year) %>%
  filter(range >= 10)

# Merge the date range information into data frame 
date_data_long <- inner_join(all_data_long, all_data_daterange)

# Find indicators where year is within 2 years of most recent GDP reading, and there is non-zero standard deviation, and there is more than 10 observations
good_indicators <- date_data_long %>% 
  group_by(countrycode, indicatorcode) %>%
  filter(max(year) >= max(max_year)-2) %>%
  filter(n() >= 10, sd(value) > 0) %>%
  filter(year <= max_year, year >= min_year)

# Find indicators that a lot of countries have data for
high_freq_indicators <- good_indicators %>%
  ungroup() %>%
  select(countrycode, indicatorcode) %>% unique() %>%
  group_by(indicatorcode) %>%
  tally() 

qplot(high_freq_indicators$n, geom="density") + 
  xlab("Number of countries with data about this indicator")

# Find indicators with support in lots of countries
high_freq_indicators_list <- (high_freq_indicators %>% filter(n > 150))$indicatorcode

great_indicators <- good_indicators %>%
  filter(indicatorcode %in% high_freq_indicators_list)

high_freq_countries <- great_indicators %>% select(countrycode, indicatorcode) %>% unique() %>% group_by(countrycode) %>% tally() 

qplot(high_freq_countries$n, geom="density") + 
  xlab("Number of indicators each country has data for")

high_freq_countries_list <- (high_freq_countries %>% filter(n > 300))$countrycode

great_indicators_and_countries <- great_indicators %>%
  filter(countrycode %in% high_freq_countries_list)

# Transform into matrix
all_data_spread_matrix <- great_indicators_and_countries %>%
  spread(key = indicatorcode, value = value) %>%
  select(-c(2:5))

# correlate
all_countries_cor_matrix <- all_data_spread_matrix %>%
  group_by(countrycode) %>%
  do(cor(x = as.matrix(.[,gdp]), y = as.matrix(select(., -matches(gdp), -countrycode)), use = "pairwise.complete", method = "spearman") %>% as.data.frame())

# Turn into matrix for pheatmap
rownames(all_countries_cor_matrix) <- all_countries_cor_matrix$countrycode
all_countries_cor_heatmap <- all_countries_cor_matrix %>% ungroup %>% select(-countrycode) %>% data.matrix()
```

## Plot a heatmap

```{r, fig.width=15, fig.height=15}
pheatmap(all_countries_cor_heatmap, cluster_rows = T, cluster_cols = T, 
         color = colorRampPalette(c("#2c7bb6", "#ffffbf", "#d7191c"))(21))
```


